---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# betterChromVAR

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/plger/betterChromVAR)](https://github.com/plger/betterChromVAR/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/plger/betterChromVAR)](https://github.com/plger/betterChromVAR/pulls)
[![Bioc release status](http://www.bioconductor.org/shields/build/release/bioc/betterChromVAR.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/betterChromVAR)
[![Bioc devel status](http://www.bioconductor.org/shields/build/devel/bioc/betterChromVAR.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/betterChromVAR)
[![Bioc downloads rank](https://bioconductor.org/shields/downloads/release/betterChromVAR.svg)](http://bioconductor.org/packages/stats/bioc/betterChromVAR/)
[![Bioc support](https://bioconductor.org/shields/posts/betterChromVAR.svg)](https://support.bioconductor.org/tag/betterChromVAR)
[![Bioc history](https://bioconductor.org/shields/years-in-bioc/betterChromVAR.svg)](https://bioconductor.org/packages/release/bioc/html/betterChromVAR.html#since)
[![Bioc last commit](https://bioconductor.org/shields/lastcommit/devel/bioc/betterChromVAR.svg)](http://bioconductor.org/checkResults/devel/bioc-LATEST/betterChromVAR/)
[![Bioc dependencies](https://bioconductor.org/shields/dependencies/release/betterChromVAR.svg)](https://bioconductor.org/packages/release/bioc/html/betterChromVAR.html#since)
[![check-bioc](https://github.com/plger/betterChromVAR/actions/workflows/check-bioc.yml/badge.svg)](https://github.com/plger/betterChromVAR/actions/workflows/check-bioc.yml)
[![Codecov test coverage](https://codecov.io/gh/plger/betterChromVAR/graph/badge.svg)](https://app.codecov.io/gh/plger/betterChromVAR)
<!-- badges: end -->

`betterChromVAR` is a much faster implementation of the popular chromVAR (Chromatin Variation Across Regions) method, used to infer TF activity from (bulk or single-cell) ATAC-seq data and motif annotations. The package also includes an ATAC-seq normalization method based on the general chromVAR logic.

## Installation instructions
<!--
Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `betterChromVAR` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("betterChromVAR")
```
-->

Install the development version from [GitHub](https://github.com/plger/betterChromVAR) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("plger/betterChromVAR")
```


