---
title: "Introduction to betterChromVAR"
author: 
  - name: Pierre-Luc Germain
    affiliation:
    - ETH Zurich, Switzerland
    email: pierre-luc.germain@hest.ethz.ch
date: "`r doc_date()`"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
bibliography: ../inst/REFERENCES.bib
package: "`r pkg_ver('betterChromVAR')`"
vignette: >
  %\VignetteIndexEntry{Introduction to betterChromVAR}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
---

# Introduction

The `r Biocpkg("chromVAR")` R package was originally developed by Alicia
Schep and colleagues from the Greenleaf lab [@schep2017chromvar].
Although originally designed for single-cell ATAC-seq data, it has been
shown to be highly sensitive for bulk data as well
[@gerbaldo2024identification]. The aim of the method is to infer, based
on the accessibility of motif matches, the relative activity of
transcription factors (TFs) in each sample or cell, adjusting for
technical biases (GC content and enrichment bias). It is recommended
that you read the `r Biocpkg("chromVAR")` documentation before using
this package.

`r Biocpkg("betterChromVAR")` is first and foremost a considerably
faster, analytical re-implementation of the original method (it is also
considerably faster than the C++ reimplementation in ArchR
[@granja2021archr]. Contrarily to the original chromVAR, it is entirely
deterministic, and achieves much higher efficiency by replacing permutations 
with an analytical solution and computing expectations and variance at the 
level of bias bins, instead of in the peak-space.

In addition, `r Biocpkg("betterChromVAR")` includes a few additions,
such as simpler weighted expectations, bias shrinkage, and an ATAC-seq
normalization method based on the chromVAR logic. Importantly, however,
not all functionalities of chromVAR have been reimplemented here:
`r Biocpkg("betterChromVAR")` chiefly focuses on the key task of
efficiently computing motif deviations.

# Installing `betterChromVAR`

`r Biocpkg("betterChromVAR")` is a `R` package available via the
[Bioconductor](http://bioconductor.org) repository. It can be installed
using the following commands in your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }

BiocManager::install("betterChromVAR")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

# Computing motif deviations with betterChromVAR

`r Biocpkg("betterChromVAR")` takes two primary inputs: 1) the counts in
peaks across cells or samples, and 2) an annotation of which TFs/motifs
match which peak[^1]. The peak counts should be provided as a
`RangedSummarizedExperiment` object[^2], while the annotation can either
be in that format too or provided as a (sparse) matrix.

[^1]: The annotation can be binary or probabilistic (i.e. from 0 to 1).
    If using motif matches, however, it is recommended to input binary
    matches, as the magnitude of motif scores is generally poorly
    correlated to actual binding.

[^2]: See `r Biocpkg("SummarizedExperiment")` if you're not familiar
    with those

There are multiple ways of generating your peak counts; the original
`r Biocpkg("chromVAR")` package includes such a function
(`getCounts()`), and the `r Githubpkg("ETHZ-INS/epiwraps", "epiwraps")`
package has some with more options (functions `peakCountsFromBAM()` and
`peakCountsFromFrags()`). Similarly, the `r Biocpkg("motifmatchr")`
package can be used to generate the motif matching annotation. An
important consideration is that the peaks or regions used for the
purpose of this analysis should have similar widths. It is thus highly
recommended, before generating the count and annotation matrices, to
resize your regions (e.g. using
`peaks <- resize(peaks, width=300, fix="center")`). The exact size can
be something not too large (otherwise the presence/absence of a motif
becomes meaningless) and ideally close to the median size of your
original peaks. In addition, is it advisable to restrict your peaks to
those that are on standard chromosomes[^3].

[^3]: See `keepStandardChromosomes()` from the
    `r Biocpkg("GenomeInfoDb")` package

Here we'll use dummy data as example:

```{r loadData}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(betterChromVAR)
})
attach(getDummyData())
counts
head(motifMatches)
```

In this case, we can see that `rowData(counts)` already has a `bias`
column indicating the GC content of the regions:

```{r inspectRowData}
rowData(counts)
```

Had this not been the case, we would first need to add this using:

```{r addGCBias, eval=FALSE}
# not run
counts <- addGCBias(counts, genome=my_genome)
```

where `my_genome` is a `r Biocpkg("BSgenome")` object or similar (e.g.
an `FaFile`[^4]).

[^4]: See the `r Biocpkg("Rsamtools")` package.

Once we have this, we can launch the computation of the deviations :

```{r betterChromVAR}
dev <- betterChromVAR(counts, motifMatches)
dev
```

The resulting `dev` object is a SummarizedExperiment with the same
columns (and colData) as the original `counts` object, but with the
motifs as rows (instead of the original peaks). The values can be
interpreted as the relative activity, across samples or cells, of the
corresponding TFs[^5]. When the function is used with default arguments,
the results are virtually the same as the original chromVAR (averaging
over the noise coming from the random background selection of the
original method).

[^5]: Note, however, that similar motifs will be given similar activity
    estimates, so that it is often hard to know which of a set of
    highly-similar motifs is in fact responsible for the observed
    signal.

The object contains two assays: the `deviations` assay contains the
bias-adjusted deviations from the expectation (by default, the average),
i.e. the difference to the expectation divided by the expectation, and
the `z` assay contains z-scores, i.e. the difference to the expectation
divided by the variance of the expectation.

If your samples/cells have similar library sizes, it is recommended that
you use the `z` assay for downstream analysis (such as differential TF
activity using `limma`). If the samples have very different library
sizes, the `z` scores will be influenced by that, and it might be
preferable to use the `deviations` assay.

The variability of each motif across the dataset is stored in the
rowData of the object :

```{r variability}
rowData(dev)
```

Note that the confidence interval is based on analytic estimates; if bootstrap 
confidence intervals (as in the original approach) are needed, see the 
`computeMotifVariability()` function.

## Individual steps of betterChromVAR

The `betterChromVAR()` function is actually a wrapper around three
steps, which can also be executed individually for more customization,
or to avoid repeating some computation multiple times.

The first step is creating the bias bins and their pairwise sampling
probabilities. This is achieved with:

```{r getBackgroundBins}
bg <- getBackgroundBins(counts)
bg
```

The second step is computing the per-bin background expectations and
variances for each sample or cell:

```{r computeBackgrounds}
bg <- computeBackgrounds(counts, bg)
bg
```

The `bg` object has now been filled with the additional information. It
can now be used for:

```{r computeDeviations}
dev <- computeDeviationsAnalytic(counts, background=bg, annotations=motifMatches)
dev
```

This last call could be made with different annotations, without needing
to recompute the previous steps.

## Including fragment length bias

In addition to the enrichment and GC bias taken into account by the
original `r Biocpkg("chromVAR")`, `r Biocpkg("betterChromVAR")` supports
an optional third dimension: fragment length bias (see
`?getBackgroundBins()` for more information). This requires the
compilation of an additional bias component (the log10-transformed mean
or median fragment length per region). If this is desired, on option is to use 
the counting functions from the
`r Githubpkg("ETHZ-INS/epiwraps", "epiwraps")` package, which can
provide this information.

## Note on single-cell data

If you are using betterChromVAR (or ChromVAR, for that matter) on
single-cell data with multiple cell types, the way to best run it
depends on whether your are interested in differences within cell types,
or between cell types. If interested in differences between cell types,
run it on the entire dataset, specifying the cell types as `grouping`
argument to `betterChromVAR()`. In this way, rare and abundant cell
types will be given the same weight in computing the expectation.

If you are interested in differences within cell types (e.g. between
samples/conditions), you should instead run the method separately for
each cell type. In this way, the background bins will be defined based
on the enrichment in the given cell type, leading to better capture of
the bias. The drawback, however, is that the values won't be comparable
across cell types. To test across conditions, we also recommend using it
on pseudobulk data.

### Execution on Atlas-scale datasets

The only step of the process that needs to be done across the entire
dataset is the computing of the expectation, i.e. the mean (or weighted
mean) for each region across all cells, and the creation of the
background bins, which depends on it. Everything else can easily be
executed in chunks (of cells), as is done for multi-threading in the
`betterChromVAR()` function.

For example, only these functions need to be called on the full dataset
:

```{r getExpectations}
ex <- getExpectation(counts)
bg <- getBackgroundBins(ex, bias=rowData(counts)$bias)
# this is equivalent to bg <- getBackgroundBins(counts)
```

Then we can apply the next steps only on subset of the data:

```{r chunkCompute}
bg2 <- computeBackgrounds(counts[,1:3], bg, expectation = ex)
dev2 <- computeDeviationsAnalytic(counts[,1:3], bg2, motifMatches)
# this should be identical to what we had run on the whole object:
identical(assay(dev)[,1:3], assay(dev2))
```

If data is on disk, rather than in memory (and you want it to stay that
way), make sure to use this procedure. In recent versions,
`getExpectation()` should use chunk-based processing. However it is
quite likely that better performance for this step would be obtained
using for instance using the `aggregateAcrossCells()` function of the
`r Biocpkg("scrapper")` package.

<br/><br/>

# Bias-normalization of bulk ATAC-seq data

The general chromVAR approach, and in particular the deterministic
version implemented here, is also amenable to be used to normalize GC-
and enrichment bias out of bulk ATAC-seq data. This is implemented in
the `CVnorm()` function. In addition, if given a grouping of the samples
(e.g. experimental conditions), the function applies a variance-based
smoothing of the bias correction, inspired by smoothed quantile
normalization (see the `r Biocpkg("qsmooth")` package or
@hicks2018smooth ). In a nutshell, if the bias in a certain background
bin is explained by experimental groups, it will be less corrected than
if it varies across samples of the groups[^6].

[^6]: Note that this is different from the original
    `r Biocpkg("qsmooth")` approach -- see `?CVnorm` for more detail.

Example usage:

```{r CVnorm}
# we assign arbitrary groups to the samples:
counts$group <- rep(LETTERS[1:2], each=5)
# we run the smoothed CVnorm:
counts <- CVnorm(counts, grouping=counts$group)
# (the normal CVnorm could be run by omitting the grouping)
counts
```

This adds a `corrected` assay to the object. Note that although the
assay is corrected for enrichment- and GC-bias (and fragment length bias if
that was provided), it is not corrected for library size differences. Rather, 
it is on the original count scale (although not integer anymore), so that it is 
amenable to use in downstream count-based analysis methods such as 
`r Biocpkg("edgeR")`.

# betterChromVAR with kNN-based background

The [scPRINTER](https://github.com/buenrostrolab/scPrinter) python
package developed by Ruochi Zhang in the Buenrostro lab includes a
GPU-based reimplementation of \textit{chromVAR}, with an important
modification: rather than cutting the bias space into bins, a
continuous, multidimensional bias space space is created from which $k$
nearest neighbors (kNN) are selected as background. This approach scales
nicely to the inclusion of more dimensions of bias. In *scPRINTER*, the 
deviations and variance across the $k$ neighbors is then computed through
permutations, as in the classical \textit{chromVAR} (although sped up by
the GPU). 

If a more complex multi-dimensional bias needs to be taken into account, 
`r Biocpkg("betterChromVAR")` implements a similar algorithm that 
combines our analytical framework with the kNN-based approach. This can for 
instance be done in the following fashion:

```{r}
bg <- getBackgroundKNN(counts)
dev <- computeDeviationsFromKNN(counts, bg, motifMatches)
```

We note, however, that this approach is considerably slower (although still 
substantially faster than the original chromVAR or the ArchR re-implementation),
requires more memory, and, at the single-cell level, produces more variable 
deviations across replicate cells.
We therefore recommend using the more scalable bin-based variant, unless 
working with bulk data with fragment length bias (or any other extra bias 
dimensions, see the `bias` argument of `getBackgroundKNN()`).

# Appendix

## References

::: {#refs}
:::

## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
